A Gentle Introduction to Structural Causal Models

### 2021-05-20 draft.Rmd
draft
presentation

The potential outcomes are conditionally independent of treatment given some set of deconfounding variables. [@hardtrecht]
(perfect RCT)

@@

(a) possible to undertake local intervention -> change f(x), regardless of f(y|x)
(b) these components are autonmous objects -> set of autonmous equations

@@

- dependecy only via vertice connection (noise)
- condition on noise, variables become independent  
- Independence of mechanisms 
- conditional independence: 
- joint distribution

@@

>>>>

The potential outcomes are conditionally independent of treatment given some set of de-confounding variables. [@hardtrecht]
(perfect RCT)

@@

(a) possible to undertake local intervention -> change f(x), regardless of f(y|x)
(b) these components are autonomous objects -> set of autonomous equations

@@

- dependency only via vertice connection (noise)
- condition on noise, variables become independent  
- Independence of mechanisms 
- conditional independence: 
- joint distribution

@@

- no need functional dependency -> less assumptions about form
- show collider variables
- iv backdoor, estimation by mechanism
[@morgan_winship_2014]

### 2021-05-20 presentation_video.Rmd
presentation

# Assumptions (1)

@@

# Assumptions (2)

@@

# Assumptions (3)

First, there was no sign to express the assignment equation and people used the '=' and one would e.g. write $A=B$.
Treating an equation as a **algebraic equation** led to confusion because those have no causal information.
This algebraic equation would imply that $B=A$ because the order has no concrete meaning in algebraic equations.
The problem is that the equation is symmetric.
The initial $'='$ sign was replaced with the ':=' which is asymmetric (Pearl, 2009) and called an **assignment**.
This misconception has caused a lot of challenges.^[For more information see @pearl2009causality]
As mentioned, we define variables as functions e.g. $A=f_A(B,U_A)$.

# Assumptions (4)

**Independence**

*Definition:*

@@

# Comparative Causal Tools

# Historical Development

* Path Analysis -> SEM -> SCM

@@

 |Method         | CBN |  SCM
|----------------|---------------------------|------------------------|
| Prediction     | $\boldsymbol{\cdot}$ Unstable                                                                                        | $\boldsymbol{\cdot}$ Stable                 
|                | $\boldsymbol{\cdot}$ Volatile to parameter changes                                                                   | $\boldsymbol{\cdot}$ More Natural Specification 
|                | $\boldsymbol{\cdot}$ Re-Estimate entire model                                                                        | $\boldsymbol{\cdot}$ Only estimate $\Delta$ CM
|                |                                                                                                                      | 
| Intervention   | $\boldsymbol{\cdot}$ Costly for Non-Markovian Models                                                                 | $\boldsymbol{\cdot}$ Pot. Cyclic Representation
|                | $\boldsymbol{\cdot}$ Unstable(Nature CP)                                                                             | $\boldsymbol{\cdot}$ Stable(Nature Eq.)
|                | $\boldsymbol{\cdot}$ Only generic estimates($\Delta$ CP)                                                             | $\boldsymbol{\cdot}$ Context specific(Invariance of Eq.)
|                |                                                                                                                      | 
|Counterfactuals | $\boldsymbol{\cdot}$ **Impossible**                                                                                  | $\boldsymbol{\cdot}$ Possible
|                | $\boldsymbol{\cdot}$ no information on latent factors($\epsilon$)                                                    | $\boldsymbol{\cdot}$ Inclusion of latent factors

@@

 -ML, BN and regression are at the lowest level in the causal hierarchy
- Prediction methods demand the least information and depend on association alone
- Associational methods ignore external changes outside of our data.
- intervention distribution has information on these external changes.
- intervention distribution is only defined in high level causal methods.

@@

The second query deals with interventions.
Here we can use @pearl2009causality do-calculus.
The do-calculus enables us to study the manipulation of parent nodes.
 There are various types of intervention.
One example is **atomic intervention,** where we set a variable to a constant.
In **policy intervention** we specify a different function for an equation.
off-policy intervention models different intervention that is not in our historical data [@oberst2019counterfactual].
Causal bayesian networks , Markov Decision Processes (MDP) and reinforcement learning model intervention.

@@

# Graphical Illustration

@@

@@

- Time in Physical Sciences: Mechanical and exact 
- Time in Social Sciences: Often Vague 
- Regular Time Specification is also more vague 
- To accomodate that issue, research on differential equation based SCMs started

>>>>

# Foundations of SCMs (1)

@@

# SCM vs SEM

@@

# Assumptions

- First, there was no sign to express the assignment equation and people used the '=' and one would e.g. write $A=B$.
- Treating an equation as a **algebraic equation** led to confusion because those have no causal information.
- This algebraic equation would imply that $B=A$ because the order has no concrete meaning in algebraic equations.
- The problem is that the equation is symmetric.
- The initial $'='$ sign was replaced with the ':=' which is asymmetric (Pearl, 2009) and called an **assignment**.
- This misconception has caused a lot of challenges.^[For more information see @pearl2009causality]
- As mentioned, we define variables as functions e.g. $A=f_A(B,U_A)$.

# Assumptions in Causality

**Established Conditions:**

Definition: **SUTVA**

'The treatment that one unit receives does not change the effect of treatment for any other unit.'[@hardtrecht]

Definition: **Consistency**

The outcome Y agrees with the potential outcome corresponding to the treatment indicator.' [@hardtrecht]

Definition: **Ignorability**

The potential outcomes are conditionally independent of treatment given some set of de-confounding variables. [@hardtrecht]
(perfect RCT)

- First tow hold for SCM counterfactuals 
- third not testable but can check via backdoor criterion in SCM

@@

@@

\footnotesize

 |Method         | CBN |  SCM
|----------------|---------------------------|------------------------|
| Prediction     | $\boldsymbol{\cdot}$ Unstable                                                                                        | $\boldsymbol{\cdot}$ Stable                 
|                | $\boldsymbol{\cdot}$ Volatile to parameter changes                                                                   | $\boldsymbol{\cdot}$ More Natural Specification 
|                | $\boldsymbol{\cdot}$ Re-Estimate entire model                                                                        | $\boldsymbol{\cdot}$ Only estimate $\Delta$ CM
|                |                                                                                                                      | 
| Intervention   | $\boldsymbol{\cdot}$ Costly for Non-Markovian Models                                                                 | $\boldsymbol{\cdot}$ Pot. Cyclic Representation
|                | $\boldsymbol{\cdot}$ Unstable(Nature CP)                                                                             | $\boldsymbol{\cdot}$ Stable
|                | $\boldsymbol{\cdot}$ Only generic estimates($\Delta$ CP)                                                             | $\boldsymbol{\cdot}$ Context specific 
|                |                                                                                                                      | 
|Counterfactuals | $\boldsymbol{\cdot}$ **Impossible**                                                                                  | $\boldsymbol{\cdot}$ Possible
|                | $\boldsymbol{\cdot}$ no information on latent factors($\epsilon$)                                                    | $\boldsymbol{\cdot}$ Inclusion of latent factors

@@

- ML, BN and regression are at the lowest level in the causal hierarchy
- Prediction methods demand the least information and depend on association alone
- Associational methods ignore external changes outside of our data.
- intervention distribution has information on these external changes.
- intervention distribution is only defined in high level causal methods.

@@

- The second query deals with interventions
- Here we can use do-calculus
- The do-calculus enables us to study the manipulation of parent nodes
- There are various types of intervention
- One example is **atomic intervention,** where we set a variable to a constant
- In **policy intervention** we specify a different function for an equation
- Off-policy intervention models different intervention that is not in our historical data
- CBN , MDP and reinforcement learning model intervention.

@@

# Graphical Tools

- Nodes 
- Edges
- Parents/Ancestors
-

# Graphical Illustration - Probabilisitic Model

@@

# Graphical Illustration - Structural Causal Model

@@

- Time in Physical Sciences: Mechanical and exact 
- Time in Social Sciences: Often Vague 
- Regular Time Specification is also more vague 
- To accommodate that issue, research on differential equation based SCMs started

# Graphical Overview

\tiny
$$
\begin{array}{|l|l|l|l|l|l|}
\hline \text { model } & \begin{array}{l}
\text { predict in IID } \\
\text { setting }
\end{array} & \begin{array}{l}
\text { predict under } \\
\text { changing } \\
\text { distributions / } \\
\text { interventions }
\end{array} & \begin{array}{l}
\text { answer } \\
\text { counter- } \\
\text { factual } \\
\text { questions }
\end{array} & \begin{array}{l}
\text { obtain } \\
\text { physical } \\
\text { insight }
\end{array} & \begin{array}{l}
\text { automatically } \\
\text { learn from } \\
\text { data }
\end{array} \\
\hline \begin{array}{l}
\text { mechanistic } \\
\text { model } \\
\end{array} & \mathrm{Y} & \mathrm{Y} & \mathrm{Y} & \mathrm{Y} & ? \\
\hline \begin{array}{l}
\text { structural } \\
\text { causal model }
\end{array} & \mathrm{Y} & \mathrm{Y} & \mathrm{Y} & \mathrm{N} & \mathrm{Y} ? ? \\
\hline \begin{array}{l}
\text { causal } \\
\text { graphical } \\
\text { model }
\end{array} & \mathrm{Y} & \mathrm{Y} & \mathrm{N} & \mathrm{N} & \mathrm{Y} ? \\
\hline \begin{array}{l}
\text { statistical } \\
\text { model }
\end{array} & \mathrm{Y} & \mathrm{N} & \mathrm{N} & \mathrm{N} & \mathrm{Y} \\
\hline
\end{array}
$$
\normalsize
